---
title: "Comparing samples visually with betta"
author: "Amy Willis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{betta-figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  `betta` is useful for formally testing differences between communities with respect to their alpha diversity. However, before ever doing any inferential test, it's best to try to visualise the difference that you are looking for. Here is a example to show you how to do that using `betta_pic`. 


```{r}
library(breakaway)
```


```{r}

frequencytablelist <- lapply(apply(toy_otu_table, 2, table), as.data.frame)
frequencytablelist <- lapply(frequencytablelist, function(x) x[x[,1]!=0,])

ob_results <- lapply(frequencytablelist[c(6:9)], objective_bayes_negbin, answers = T, plot=F, output = F)

lower <- unlist(lapply(ob_results, function(x) x[["ci"]][2]))
upper <- unlist(lapply(ob_results, function(x) x[["ci"]][1]))
means <- unlist(lapply(ob_results, function(x) x[["est"]]))
means
standard_deviations <- unlist(lapply(ob_results, function(x) x[["semeanest"]]))

## Find how many otus are in the cyanobacteria genus
cyano_nodes <- apply(toy_otu_table[grepl("Cyano", toy_taxonomy), ], 2, function(x) sum(x>0))
cyano_nodes <- cyano_nodes

bloom1 <- toy_metadata[, "bloom2"][6:9]
bloom <- bloom1 == "yes"
```

The following is the default option for plotting. It shows the mean estimates with lines up to +/-2  standard deviations. In this way, it is a visual mimic of the procedure that underpins `betta`. Feel free to pull apart the source code  to optimise it for your purposes. It uses base graphics.

```{r, fig.show='hold'}
betta_pic(means, standard_deviations, mypch = ifelse(bloom, 16, 1))
```

The following is a nice option if you would like to use ggplot. 

```{r, fig.show='hold'}
library(ggplot2)
plot_data_frame <- data.frame("index" = 1:4, lower, upper, means, standard_deviations, "bloom"=bloom)
plot_data_frame

ggplot(plot_data_frame, aes(x = index, y = means, col = bloom)) +
  geom_segment(aes(x = index, xend = index, y = lower, yend = upper, col = bloom), lwd = 1.25) +
  geom_point(aes(x = index, y = means)) + xlab("Samples") + ylab("Richness estimates (Objective Bayes)")
```