---
title: "Getting started with breakaway"
author: "Amy Willis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`breakaway` is a package for species richness estimation and modelling. As the package has grown and users have requested more functionality, it contains some basic generalisations of the statistical philosophy that underpins `breakaway` to the general alpha diversity case. Because of the flexibility of the modelling strategies, most users of breakaway are microbial ecologists with very large OTU tables, however, nothing should exclude a macroecologist from using the same tools. If you have a macroecology dataset and want to use this package, I would love to hear from you so please feel free to contact me (email or via Github's Issues/Projects infrastructure).

## Vignette Info

This vignette will lead you through the most basic way to use `breakaway`. For a more in depth discussion of how and why estimating species richness is possible, check out the *Introduction to diversity estimation* vignette. 

## Diving in!

Download the latest version of the package from github.

```{r}
library(breakaway)
data(toy_otu_table)
## For historical reasons we're going to rename it:
otu_data <- toy_otu_table
```

## Creating frequency tables

We're now going to "collapse" the otu_data's columns (samples) into frequency tables. Frequency tables...

```{r}
frequencytablelist <- build_frequency_count_tables(otu_data)
head(frequencytablelist[[63]])
```

Interpretation: In this sample, there were `r frequencytablelist[[63]][frequencytablelist[[63]][, 1]==1, 2]` different species observed only once (singletons), `r frequencytablelist[[63]][frequencytablelist[[63]][, 1]==2, 2]` different species observed only twice, ..., `r tail(frequencytablelist[[63]], 1)[2]` species observed `r tail(frequencytablelist[[63]], 1)[1]` times.


## Estimating species richness

Let's run breakaway on the first frequency count table

```{r}
breakaway(frequencytablelist[[1]])
```

You should get some output to screen, including your estimate & s.e. You can also investigate a plot of the fits to the ratios as follows: 

```{r}
plot(breakaway(frequencytablelist[[1]]))
``` 

Note that it is not a fit to the frequencies, it is a fit to the ratios of frequencies. You would never need to include this type of plot in one of your papers. It is solely for you to check for model misspecification. What's model misspecification? If the model fit (pink circles) don't remotely follow the pattern of the observed ratios (green triangles), that's model misspecification.

Sometimes, breakaway's usual procedure doesn't work, that is, it gives a negative estimate, which is of course silly. In that case, breakaway returns a different model's result. It's called the \code{WLRM}. There isn't a picture. Here is an example of a case where breakaway returns the \code{WLRM}.

```{r}
breakaway(frequencytablelist[[2]])
```
breakaway can defer to the \code{WLRM} for several reasons. Perhaps there are too many singletons. Perhaps there isn't a long enough tail. Perhaps there is false diversity. Regardless, we can see if this failure was sensitive to the singleton count by running `breakaway_nof1()`. This requires no singleton count (implicit is that the singleton count was erroneous) and predicts it from the other frequencies: 

```{r}
breakaway_nof1(frequencytablelist[[2]][-1,])
```

The reference for `breakaway_nof1()` is:
Willis, A. (2016). Species richness estimation with high diversity but spurious singletons.

`breakaway_nof1()` is an exploratory tool for assessing sensitivity of
breakaway to the singleton count. I recommend it as a sensitivity analysis
rather than for diversity
estimation. 


## Next steps 

- The above discussion focused exclusively on richness. Check out `github.com/adw96/DivNet` for alpha and beta diversity tutorials!
- To learn about modelling alpha diversity, see the vignette *Introduction to hypothesis testing for diversity*
- To hear more details about how species richness estimation works, check out *Introduction to diversity estimation*

