---
title: "Introduction to hypothesis testing for diversity"
author: "Amy Willis"
date: "`r Sys.Date()`"
output: 
  output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{intro-hypothesis-testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE) 
```
  
This tutorial will talk through hypothesis testing for alpha diversity indices using the functions `betta` and `betta_random`. 

## Disclaimer

*Disclaimer*: If you have not taken a introductory statistics class or devoted serious time to learning introductory statistics, I strongly encourage you to reconsider doing so before ever quoting a p-value or doing modeling of any kind. An introductory statistics class will teach you valuable skills that will serve you well throughout your entire scientific career, including the use and abuse of p-values in science, how to responsibly fit models and test null hypotheses, and an appreciation for how easy it is to inflate the statistical significance of a result. Please equip yourself with the statistical skills and scepticism necessary to responsibly test and discuss null hypothesis significance testing.

## Preliminaries

Download the latest version of the package from github.

```{r, include=FALSE}
# install.packages("devtools")
# devtools::install_github("adw96/breakaway")
library(tidyverse)
library(breakaway)
library(magrittr)
library(phyloseq)
```

Let's use the Whitman et al dataset from `corncob` as our example.

```{r}
# devtools::install_github("bryandmartin/corncob")
library(corncob)
data("soil_phylo")
soil_phylo %>% sample_data %>% head
```

I'm only going to consider samples amended with biochar, and I want to look at the effect of `Day`. This will tell us how much diversity in the soil changed from Day 0 to Day 82. (Just to be confusing, Day 82 is called Day 2 in the dataset.)

```{r}
subset_soil <- soil_phylo %>%
  subset_samples(Amdmt == 1) %>% # only biochar
  subset_samples(Day %in% c(0, 2))  # only Days 0 and 82
```

I now run breakaway on these samples to get richness estimates, and plot them.

```{r}
richness_soil <- subset_soil %>% breakaway
plot(richness_soil, physeq=subset_soil, color="Day", shape = "ID")
```

Don't freak out! Those are wide error bars, but nothing went wrong -- it's just really hard to estimate the true number of unknown species in soil. `breakaway` was developed to deal with this, and to make sure that we account for that uncertainty when we do inference.

We can get a table of the estimates and their uncertainties as follows:

```{r}
summary(richness_soil) %>% as_tibble
```

If you haven't seen a `tibble` before, it's like a `data.frame`, but way better. Already we can see that we only have 10 rows printed as opposed to the usual bagillion.

## Inference

The first step to doing inference is to decide on your design matrix. We need to grab our covariates into a data frame (or tibble), so let's start by doing that:

```{r}
meta <- subset_soil %>%
  sample_data %>%
  as_tibble %>%
  mutate("sample_names" = subset_soil %>% sample_names )
```

That warning is not a problem -- it's just telling us that it's not a phyloseq object anymore.

Suppose we want to fit the model with Day as a fixed effect. Here's how we do that,

```{r}
combined_richness <- meta %>%
  left_join(summary(richness_soil),
            by = "sample_names")
# Old way (still works)
bt_day_fixed <- betta(chats = combined_richness$estimate,
                      ses = combined_richness$error,
                      X = model.matrix(~Day, data = combined_richness))
# Fancy new way -- thanks to Sarah Teichman for implementing!
bt_day_fixed <- betta(formula = estimate ~ Day, 
                      ses = error, data = combined_richness)
bt_day_fixed$table
```

So we see an estimated increase in richness after 82 days of 122 taxa, with the standard error in the estimate of 171. A hypothesis test for a change in richness would *not be rejected* at any reasonable cut-off (p = 0.475).

Alternatively, we could fit the model with plot ID as a random effect. Here's how we do that:

```{r}
# Old way (still works)
bt_day_fixed_id_random <- betta_random(chats = combined_richness$estimate,
                                       ses = combined_richness$error,
                                       X = model.matrix(~Day, data = combined_richness),
                                       groups=combined_richness$ID)
# Fancy new way 
bt_day_fixed_id_random <-
  betta_random(formula = estimate ~ Day | ID, 
               ses = error,  data = combined_richness)
bt_day_fixed_id_random$table
```

Under this different model, we see an estimated increase in richness after 82 days of 258 taxa, with the standard error in the estimate of 161. A hypothesis test for a change in richness still would *not be rejected* at any reasonable cut-off (p = 0.11).

If you choose to use the old way, the structure of `betta_random` is to input your design matrix as `X`, and your random effects as `groups`, where the latter is a categorical variable. Otherwise, the input looks like how you would hand this off to a regular mixed effects model in the package `lme4`!

# Using `betta` with DivNet

Maybe you don't care about richness... but you care about Shannon or Simpson diversity! [`DivNet`](https://github.com/adw96/DivNet) is our `R` package for estimating Shannon and Simpson diversity.

```{r, include=FALSE}
devtools::install_github("adw96/DivNet")
library(DivNet)
```

DivNet can be slow when you have a large number of taxa (but we are working on it!), so to illustrate we are going to estimate phylum-level Shannon diversity:

```{r}
soil_phylum <- subset_soil %>%
  tax_glom(taxrank="Phylum")
```

Easter egg: `phyloseq::tax_glom` can be incredibly slow! [Mike McLaren](https://github.com/mikemc) is a total champ and rewrote it faster -- check out his package [`speedyseq`](https://github.com/mikemc/speedyseq) and `speedyseq::tax_glom` in particular.

Let's treat all samples as independent observations (`X = NULL`) and fit the DivNet model:

(Check out the full documentation for details, including how to run in parallel)

```{r, include=TRUE, cache=TRUE, message=FALSE, results="hide"}
dv <- DivNet::divnet(soil_phylum, X = NULL)
```

This produces an object containing common diversity estimates:

```{r}
dv
```

We can look at the first few Shannon diversity estimates with the following:

```{r}
combined_shannon <- meta %>%
  left_join(dv$shannon %>% summary,
            by = "sample_names")
combined_shannon
```

You might notice that the estimates are not different from the plug-in estimate (only because we used `X = NULL`), but we have standard errors! That's the real advantage of using DivNet :)


```{r}
bt_day_fixed_id_random <- betta_random(formula = estimate ~ Day | ID, 
               ses = error,  data = combined_shannon)
bt_day_fixed_id_random$table
```

and similarly for no random effects.

If you are interested in generating confidence intervals for and testing hypotheses about
linear combinations of fixed effects estimated in a `betta` or `betta_random` model, we
recommend using the 'betta_lincom' function.

For example, to generate a confidence interval for $\beta_0 + \beta_1$ (i.e., intercept plus
'Day2' coefficient, or in other words, the mean Shannon diversity in soils on day 82 of the experiment) in the Shannon diversity model we fit in the previous code chunk,
we run the following code:

```{r}

betta_lincom(fitted_betta = bt_day_fixed_id_random,
             linear_com = c(1,1),
             signif_cutoff = 0.05)


```

Here, we've set the `linear_com` argument equal to `c(1,1)` to tell `betta_lincom` to construct
a confidence interval for $1 \times \beta_0 + 1 \times \beta_1$. Because we set `signif_cutoff` equal to $0.05$, `betta_lincom` returns a $95\%  = (1 - 0.05)*100\%$ confidence interval.
The p-value reported here
is for a test of the null hypothesis that $1 \times \beta_0 + 1 \times \beta_1 = 0$ -- 
unsurprisingly, this is small. (If you are confused about why this is ``unsurprising,''think about what $1 \times \beta_0 + 1 \times \beta_1 = 0$ 
represents and when a Shannon diversity can be zero).

The syntax and output using `betta_lincom` with a `betta` object as input is exactly the same 
as with a `betta_random` object, so we haven't included a separate example for this case.

And there you have it! That's how to do hypothesis testing for diversity!


If you use our tools, please don't forget to cite them!

- `breakaway`: Willis & Bunge. (2015). *Estimating diversity via frequency ratios*. Biometrics. [doi:10.1111/biom.12332](https://doi.org/10.1111/biom.12332).
- `DivNet`: Willis & Martin. (2018+). *DivNet: Estimating diversity in networked communities*. bioRxiv. [10.1101/305045](https://doi.org/10.1101/305045).
- `betta`: Willis, Bunge & Whitman. (2016). *Improved detection of changes in species richness in high diversity microbial communities*. Journal of the Royal Statistical Society: Series C. [doi:10.1111/rssc.12206](https://doi.org/10.1111/rssc.12206).
