---
title: "Introduction to hypothesis testing for diversity"
author: "Amy Willis"
date: "`r Sys.Date()`"
output: 
  output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{intro-hypothesis-testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  
This tutorial will talk through hypothesis testing for alpha diversity indices using the functions `betta` and `betta_random`. 
  
## Preliminaries
  
Download the latest version of the package from github.
  
```{r, include=FALSE}
# install.packages("devtools")
# devtools::install_github("adw96/breakaway")
library(tidyverse)
library(breakaway)
library(magrittr)
library(phyloseq)
```

Let's use the Whitman et al dataset from `corncob` as our example. 

```{r}
# devtools::install_github("bryandmartin/corncob")
library(corncob)
data("soil_phylo")
soil_phylo %>% sample_data %>% head
```

I'm only going to consider samples amended with biochar, and I want to look at the effect of `Day`. This will tell us how much diversity in the soil changed from Day 0 to Day 82. (Just to be confusing, Day 82 is called Day 2 in the dataset.) 

```{r}
subset_soil <- soil_phylo %>%
  subset_samples(Amdmt == 1) %>% # only biochar
  subset_samples(Day %in% c(0, 2))  # only Days 0 and 82
```

I now run breakaway on these samples to get richness estimates, and plot them.

```{r}
richness_soil <- subset_soil %>% breakaway
plot(richness_soil, physeq=subset_soil, color="Day", shape = "ID")
```

Don't freak out! Those are wide error bars, but nothing went wrong -- it's just really hard to estimate the true number of unknown species in soil. `breakaway` was developed to deal with this, and to make sure that we account for that uncertainty when we do inference.

We can get a table of the estimates and their uncertainties as follows:

```{r}
summary(richness_soil) %>% as_tibble
```

If you haven't seen a `tibble` before, it's like a `data.frame`, but way better. Already we can see that we only have 10 rows printed as opposed to the usual bagillion. 

## Inference

The first step to doing inference is to decide on your design matrix. We need to grab our covariates into a data frame (or tibble), so let's start by doing that:

```{r}
meta <- subset_soil %>% 
  sample_data %>%
  as_tibble %>%
  mutate("sample_names" = subset_soil %>% sample_names )
```

That warning is not a problem -- it's just telling us that it's not a phyloseq object anymore.

Suppose we want to fit the model with Day as a fixed effect. Here's how we do that,

```{r}
combined_richness <- meta %>% 
  left_join(summary(richness_soil), 
            by = "sample_names")
bt_day_fixed <- betta(chats = combined_richness$estimate,
                      ses = combined_richness$error,
                      X = model.matrix(~Day, data = combined_richness)) 
bt_day_fixed$table
```

So we see an estimated increase in richness after 82 days of 122 taxa, with the standard error in the estimate of 171. A hypothesis test for a change in richness would *not be rejected* at any reasonable cut-off (p = 0.475).

Alternatively, we could fit the model with plot ID as a random effect. Here's how we do that:

```{r}
bt_day_fixed_id_random <- betta_random(chats = combined_richness$estimate,
                                       ses = combined_richness$error,
                                       X = model.matrix(~Day, data = combined_richness),
                                       groups=combined_richness$ID) 
bt_day_fixed_id_random$table
```

Under this different model, we see an estimated increase in richness after 82 days of 234 taxa, with the standard error in the estimate of 163. A hypothesis test for a change in richness still would *not be rejected* at any reasonable cut-off (p = 0.153).

The overall structure of `betta` and `betta_random` is to input your design matrix as `X`, and your random effects as `groups`, where the latter is a categorical variable. 

A future version of breakaway hopes to make this interface easier to use a more familiar to users of `lme4` and other mixed effects models. If this is late 2020 and it's not done, please come and find me (Amy) and make me feel embarrassed. 

# Using `betta` with DivNet

Maybe you don't care about richness... but you care about Shannon or Simpson diversity! [`DivNet`](https://github.com/adw96/DivNet) is our `R` package for estimating Shannon and Simpson diversity. 

```{r, include=FALSE}
devtools::install_github("adw96/DivNet")
library(DivNet)
```

DivNet can be slow when you have a large number of taxa (but we are working on it!), so to illustrate we are going to estimate phylum-level Shannon diversity:

```{r}
soil_phylum <- subset_soil %>%
  tax_glom(taxrank="Phylum")
```

Easter egg: `phyloseq::tax_glom` can be incredibly slow! [Mike McLaren](https://github.com/mikemc) is a total champ and rewrote it faster -- check out his package [`speedyseq`](https://github.com/mikemc/speedyseq) and `speedyseq::tax_glom` in particular.

Let's treat all samples as independent observations (`X = NULL`) and fit the DivNet model:

(Check out the full documentation for details, including how to run in parallel)

```{r, include=FALSE, cache=TRUE}
dv <- DivNet::divnet(soil_phylum, X = NULL)
```

This produces an object containing common diversity estimates:

```{r}
dv
```

We can look at the first few Shannon diversity estimates with the following:

```{r}
combined_shannon <- meta %>% 
  left_join(dv$shannon %>% summary, 
            by = "sample_names")
combined_shannon
```

You might notice that the estimates are not different from the plug-in estimate (only because we used `X = NULL`), but we have standard errors! That's the real advantage of using DivNet :)


```{r}
bt_day_fixed_id_random <- betta_random(chats = combined_shannon$estimate,
                                       ses = combined_shannon$error,
                                       X = model.matrix(~Day, data = combined_shannon),
                                       groups=combined_shannon$ID) 
bt_day_fixed_id_random$table
```

and similarly for no random effects. 

And there you have it! That's how to do hypothesis testing for diversity! 